---
title: "sketch3"
author: "Caryn Johansen"
date: "1/17/2018"
output: 
  html_document: 
    keep_md: yes
---

## Introduction

Return to the abalone data set from sketch 1 to do some 3D plotting.

## Set Up

```{r}
library(tidyverse)
library(plot3D)
library(GGally)
```


## Data

```{r}
abalone <- read.delim("../sketch1/data/abalone.data", sep=",", header=F)
```

### Attribute information

| Name | Data Type | Measurement Unit | Description |
| ---- | --------- | ---------------- | ----------- |
| Sex | nominal | -- | M, F, and I (infant) |
| Length | continuous | mm | Longest shell measurement |
| Diameter | continuous | mm | perpendicular to length |
| Height | continuous | mm | with meat in shell |
| Whole weight | continuous | grams | whole abalone |
| Shucked weight | continuous | grams | weight of meat |
| Viscera weight | continuous | grams | gut weight (after bleeding) |
| Shell weight | continuous | grams | after being dried |
| Rings | integer | -- | +1.5 gives the age in years |

```{r}
colnames(abalone) <- c("sex","length", "diameter", "height", "whole", "shucked", "viscera","shell", "rings")
```

### clean data

```{r}
summary(abalone)
str(abalone)
dim(abalone)
abalone <- abalone %>% na.omit()
dim(abalone)
```

The curious thing about this data set is that some samples have a height of zero. That may be a typo, or an indication of missing data, but there is probably not a real abalone with a height of zero. So let's remove those.

```{r}
abalone %>% filter(height == 0)
```

Yeah, the lengths and diameter are middle of the road, and so the height seems to be incorrect here. REMOVE.

```{r}
abalone <- abalone %>% filter(height != 0)
```


## 3D plots

Summarizing the data to figure out what values works best on what axes

```{r}
abalone %>% gather(trait, value, -sex) %>%
  group_by(trait) %>% 
  summarise(traitMin = min(value), traitMax = max(value))
```

I'm curious about the number of rings, and I want to know what the best predictors for the number of rings are.

```{r}
abalone %>% 
  gather(trait, value, -rings) %>% 
  ggplot(aes(x=value, y=rings)) + geom_point() + facet_wrap(~trait, scale="free")
abalone %>%
  gather(trait, value, -c(sex, rings)) %>%
  ggplot(aes(x=value, y=rings, color=sex)) + geom_point() + facet_wrap(~trait, scale="free")
```

In general, it appears that the larger an abalone is, the more rings it will have.

Length, diameter and height

```{r}
points3D(abalone$height, abalone$diameter, abalone$rings, phi = 30, theta = 30)
```

So there are these two points that are way out there. What's going on with those?

```{r}
abalone %>% ggplot(aes(x=height, y=rings)) + geom_point()
```

```{r}
tail(sort(abalone$height), 2)
weird1 <- abalone %>% filter(height == tail(sort(height), 2)[1])
weird2 <- abalone %>% filter(height == tail(sort(height), 2)[2])

abalone %>% ggplot(aes(x=length, y=diameter, color=height)) + geom_point() +
  geom_point(data=weird1, color="red") +
  geom_point(data=weird2, color="red")
```

These two abarant points are more or less middle of the pack in otherwise pretty linearly correlated traits (color height is a pretty nice gradient with a peppering of light spots).

```{r}
abalone %>% ggplot(aes(x=length, y=diameter, color=height)) + geom_point() +
  geom_point(data=weird1, color="red") +
  geom_point(data=weird2)
```

Bringing those two spots to the front with the same coloring scheme. Ah, the agony of removal! I cannot say if they are wrong - maybe the middle one is a rad mutant, or actually reflects the actual biodiversity range and the people who collected it didn't like handling long 

```{r}
abalone %>% ggplot(aes(x=length, y=whole, color=height)) + geom_point() +
  geom_point(data=weird1, color="red") +
  geom_point(data=weird2)
abalone %>% ggplot(aes(x=length, y=rings, color=height)) + geom_point() +
  geom_point(data=weird1, color="red") +
  geom_point(data=weird2)
```

However, the top weight is so middle of the pack that it's hard to not think it's a data entry error.

```{r}
 tail(sort(abalone$height), 10)
abalone <- abalone %>% filter(height < 0.3)
abalone %>% ggplot(aes(x=height, y=rings)) + geom_point()
```

Ok, agony over. Back to rings.

```{r}
abalone %>%
  gather(trait, value, -c(sex, rings)) %>%
  ggplot(aes(x=value, y=rings, color=sex)) + geom_point() + facet_wrap(~trait, scale="free")
```
```{r}
points3D(abalone$height, abalone$diameter, abalone$rings, phi = 30, theta = 30)
```

## Predictive model

remove 10% of the dataset for prediction and model testing

```{r}
n_samples <- dim(abalone)[1]
test_remove <- sample(1:n_samples, 0.1*n_samples)
test_abalone <- abalone[test_remove,]
train_abalone <- abalone[-test_remove,]
```

#### Starting with the most basic

```{r}
md1 <- lm(rings ~ height, data=train_abalone)
summary(md1)
train_abalone %>% ggplot(aes(x=height, y=rings)) + geom_point() + geom_abline(intercept = coef(md1)[1], slope=coef(md1)[2])
```

That's not terrible, but it also seems not that great - it's failing to cover the upper variation of rings.

Is height or length a better predictor?

```{r}
md2 <- lm(rings ~ length, data=train_abalone)
summary(md2)
anova(md1, md2)
```

Height has a larger adjusted R squared, and has a lower RSS. I'll use height going forward as a better predictor, especially since they are very correlated

```{r}
md3 <- lm(rings ~ height + diameter, data= train_abalone)
summary(md3)
anova(md1, md3)
train_abalone %>% ggplot(aes(x=height, y=rings)) + geom_point() + geom_abline(intercept = coef(md3)[1], slope=coef(md3)[2])
```

The second model is different from the other, but not really that much better.

Diameter does have a significant effect on the number of rings, but diameter and height are also correlated, so we may be getting a confounding effect here.

### Adding Random effects

```{r}
library(lme4)
```

Adding sex identification as a random effect - each sex (M, F, I) will have a separate effect on the number of rings.
I think this will improve the prediction, as the infant abalone take up a majority of the low-ring numbers.

```{r}
md4 <- lmer(rings ~ (1|sex) + height, data=train_abalone)
summary(md4)
cat("==============\n")
coef(md4)
```

```{r}
train_abalone %>% 
  ggplot(aes(x=height, y=rings, color=sex)) + 
  geom_point() + 
  geom_smooth(method = "lm")
  
  # geom_smooth(formula = rings ~ (1|sex) + height)
```

Well... Maybe this is not a linear fit. Future models may do better with a polynomial or logarithmic fit, or thoughtful transoformations of the data.

### Prediction

Going with the random effects model to predict ring values from my testing data set

```{r}
test_abalone <- test_abalone %>% mutate(predictRings = predict(md4, .))
test_abalone %>% 
  ggplot(aes(x=rings, y=predictRings)) + geom_point() + xlim(0,30) + ylim(0,30) +
  geom_abline(intercept = 0, slope = 1, color="red")
```

I've built better models in my life! Looking forward to figuring this one out...

## What I learned

I started out today wanting to fiddle with 3D plots, but got caught up with predicting ring number. That's ok, but perhaps I should keep these sketched more focused.

I learned about theta and phi in the 3D plotting library. Tomorrow perhaps I should just focus on visualizing 3D stuff.
3D visualization can be tricky, because sometimes it adds more confusion than information. 

Today I also touched on some basics of linear modeling - comparing models and parameters, using a model to predict values, and assessing my model using a testing data set.
