---
title: "Chapter 4"
author: "Caryn Johansen"
date: "1/2/2018"
output: 
  html_document: 
    keep_md: yes
---

```{r}
library(tidyverse)
library(GGally)
```

I have very little information about the raw data. It's from "Introduction to Statistical Thinking", found at: http://www-bcf.usc.edu/~gareth/ISL/data.html

It's heart attack data, perhaps from people in South Africa.

## data

```{r}
heart <- read.delim("data/Heart.csv", header=T, sep=",",row.names = 1)
colnames(heart)
```

# Data cleaning

look for and remove missing data and re-factor the sex, heart disease T/F

I can do that with a single command:

```{r}
heart <- read.delim("data/Heart.csv", header=T, sep=",",row.names = 1)
summary(heart)
heart <- heart %>% 
  na.omit() %>% 
  mutate(AHD = as.factor(if_else(AHD == "No", 0, 1)), Ca = as.factor(Ca), Sex = as.factor(Sex))
summary(heart)
```

# Exploring data

```{r}
heart %>% 
  gather(trait, value) %>% 
  ggplot(aes(x=value)) + geom_histogram(stat="count") + facet_wrap(~trait, scales = "free")
```

I'm going to remove a few variables.

```{r}
heart <- heart %>% select(-c(Thal, Slope, RestECG, ExAng, ChestPain))
heart %>% 
  gather(trait, value) %>% 
  ggplot(aes(x=value)) + geom_histogram(stat="count") + facet_wrap(~trait, scales = "free")
```


# Logistic Regression

```{r}
heartfit <- glm(AHD ~ ., data = heart, family = "binomial")
summary(heartfit)
```

# Compounding factors

```{r}
heart %>% ggpairs(columns = c("Age", "AHD", "Ca"), aes(color=Sex))
```

A fairly incomplete look, but it's the end of the day. It will be useful to build on this later, and it seems like a good dataset to deep dive into logistic regressions.